---
title:  "Malaria Risk Prediction in Kenya using Artificial Intelligence (AI) & Machine Learning(ML) Models"
author:
- name: D.K.Muriithi
  affiliation: Chuka University, Kenya
#institute: CDAM
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.height = 4,
	fig.width = 8,
	message = FALSE,
	warning = FALSE,
	comment = NA)

```

# Get started

# Set a working directory 

 This is a default location where R looks for files and saves outputs
 
```{r}
setwd("~/2025-AMMnet_Malaria_Modeling_Workshop")

```

# Install and load necessary libraries

**Loading libraries**

```{r}
library(caret)         ## for training machine learning models
library(psych)         ## for description of  data
library(ggplot2)       ## for data visualization
library(caretEnsemble) ## enables the creation of ensemble models
library(tidyverse)     ## for data manipulation
library(mlbench)       ## for benchmarking ML Models
library(flextable)     ## to create and style tables
library(mltools)       ## for hyperparameter tuning
library(tictoc)        ## for determining the time taken for a model to run
library(ROSE)          ## for random oversampling
library(smotefamily)   ## for smote sampling
library(ROCR)          ## for ROC curve
library(pROC)          ## for visualizing, smoothing, and comparing ROC curves
library(e1071)         ## for statistical modeling and  machine learning tasks(SVM)
library(class)         ## for classification using k-Nearest Neighbors and other methods
library(caTools)       ## for splitting data into training and testing sets
library(MASS)          ## provides plotting functions and datasets
library(ISLR)          ## for practical applications of statistical learning methods
library(boot)          ## useful for performing bootstrap resampling
library(cvTools)       ## contains functions for cross-validation, bootstrapping, & other resampling methods
library(iml)           ## provide tools to analyze and interpret machine learning models
library(lime)          ## powerful tools for interpreting machine learning models
library(DALEX)         ## powerful tool for interpreting machine learning models. 
library(rio)           ## for easy data import, export(saving) and conversion
library(esquisse)      ## GUI tool that allows users to easily create ggplot2 plots interactively

```

# Load and prepare data/Exporatory of the dataset

```{r}
library(rio) ## for easy data import, export(saving) and conversion

data = import("Malaria Dataset.csv")


#head(data)          # for the 1st few rows in the dataset
#tail(data)          # for the last few rows in the dataset

```


# Exploratory Data Analysis (EDA)

Before we start visualizing our data, we need to understand the characteristics of our data. The goal is to get an idea of the data structure and to understand the relationships between variables.

Here are some functions that can help us understand the structure of our data:

```{r}
#dim(data)           # for dimensions of dataset

#summary(data)       # for summary of descriptive statistics
#describe(data)      # for descriptive statistics


```

## Check for data structure 
```{r}
str(data)     
```

## Convert all character variables to factors
```{r}
data[] <- lapply(data, function(x) if(is.character(x)) as.factor(x) else x)

```

## Diagnose the data set
```{r}
library(gtsummary)    #create publication-ready summary tables for regression models and descriptive statistics
library(flextable)    #Creates highly customizable tables suitable for reporting and publication
library(dlookr)       #Analyzes the structure and quality of the dataset

diagnose(data) |> flextable()

#Explore individual columns/variables
#unique(data$Region)                          # unique values for single column
#table(data$Malaria_Result)                           # frequency for a single column
#table(data$Region, data$Intervention_Type) # frequencies for multiple columns
```


## Check for zero variance predictors:
```{r}
nzv <- nearZeroVar(data[,-18], saveMetrics = TRUE)
print(nzv)

##The results above show that there is no feature with zero variance
## Remove nzv

#data <- data[, !nzv$nzv]
#dim(data)

```

## Visualizing the Target Variable (Malaria Test Results)
```{r, fig.height = 6}
library(ggplot2)       ## for data visualization

# Plot Target variable using ggplot2 function
# Sample dataset
dt <- data.frame(
  Malaria_Result = c("Negative", "Positive"),
  Respondent = c(820, 180)) # Replace with actual numbers

# Calculate percentages
dt1 <- dt %>%
  mutate(Percentage = Respondent / sum(Respondent) * 100)

# Create the bar plot
dt1 |> ggplot(aes(x = Malaria_Result, y = Respondent, fill = Malaria_Result)) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  geom_text(aes(label = paste0(Respondent, " (", round(Percentage, 1), "%)")),vjust = -0.5, size = 5) +
  labs(title = "Imbalance Malaria Data",
       x = "Malaria Test Result",
       y = "Respondent",
       fill = "Results") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) # Align the title to the center

```

# Data Partition for Machine Learning
```{r}
library(caret)

set.seed(123)

# Create a partition: 75% for training, 25% for testing
index <- createDataPartition(data$Malaria_Result,p = 0.75, list = FALSE)

# Create training and testing sets
train <- data[index, ]
test <- data[-index, ]

# Get the dimensions of your train and test data
dim(train)
dim(test)

```


```{r, fig.height=6}
##frequency distribution of classes of the target variable in the train dataset
#table(train$Malaria_Result)
#table(test$Malaria_Result)

# Plot Target variable using ggplot2 function
# Sample dataset
dt1 <- data.frame(
  Malaria_Result = c("Negative", "Positive"),
  Respondent = c(615, 135)) # Replace with actual numbers


# Calculate percentages
dt1 <- dt1 |>
  mutate(Percentage = Respondent / sum(Respondent) * 100)

# Create the bar plot
p1 = dt1 |> ggplot(aes(x = Malaria_Result, y = Respondent, fill = Malaria_Result)) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  geom_text(aes(label = paste0(Respondent, " (", round(Percentage, 1), "%)")),vjust = -0.5, size = 5) +
  labs(title = "Imbalance Malaria Data (Training set)",
       x = "Malaria Test Result",
       y = "Respondent",
       fill = "Results") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) # Align the title to the center
print(p1)
```

# View the models in CARET
```{r}
models= getModelInfo()
#names(models)
```


# Resampling techinque (Over-sampling)

The minority class is duplicated (synthetically) until both classes have roughly the same number of samples.

```{r, fig.height= 6}
library(ROSE)    ## used for balancing imbalanced datasets (e.g., in binary classification)

set.seed(123)    ## ensures reproducibility: Every time you run this code, you get the same random over-sampling result

over<- ovun.sample(Malaria_Result~., data = train, method = "over", N = 1230 )$data # over-samples the minority class to balance the dataset
# Calculate counts and percentages
over_summary <- over |>
  group_by(Malaria_Result) |>
  summarise(Count = n()) |>
  mutate(Percent = round(Count / sum(Count) * 100, 1))
  
  # Plot
p2 = ggplot(over, aes(x = Malaria_Result, fill = Malaria_Result)) +
  geom_bar() +
  geom_text(data = over_summary, aes(x = Malaria_Result, y = Count, 
             label = paste0(Count, "(", Percent, "%)")), 
            vjust = -0.5,     # moves text higher vertically
            hjust = 0.5,      # centers text horizontally
            size = 5) +
labs(title ="Balanced Malaria Data(Training set)",
     y = "Respondent",
     x = "Malaria Test Result") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) # Align the title to the center

print(p2)

```


```{r, fig.height=6, fig.width=12}
library(patchwork)
library(easystats)

# Display the nested plot
plots(p1, p2, n_columns = 2, tags = paste("Fig.", 1:23))
```

# Cross validation technique

**Cross-validation (CV)** is a resampling technique used to evaluate the performance of a machine learning model and test its ability to generalize to unseen data. The idea is to train and test the model on different subsets of the data to reduce bias and variance in performance estimation. It helps in detecting overfitting and selecting better hyperparameters.

  <div style='text-align:center;margin:auto;display:block:'><IMG  src=  "C:\Users\Admin\Documents\2025_CDAM_WORKSHOP_1\Bais-Variance.png"></div>
  
**Two key contributors to model performance are bias and Variance**

**Bias**: We want prediction that are close to the true value of the outcome we arE trying to predict

**Variance:** We want predictions that do not vary too much around the true value

**NB:**A model that performs well on new data will balance these two


**Hyperparameters** are the settings you specify before training a machine learning model. They control the learning process and influence how well the model performs, but they are not learned from the data.

```{r}
#Creating the train-Control scheme to avoid overfitting & underfitting

library(caret)

control <- trainControl(
  method = "repeatedcv",     # Use repeated cross-validation
  number = 10,               # 10-fold cross-validation
  repeats = 5,               # Repeat the 10-fold CV 5 times
  classProbs = TRUE,         # Compute class probabilities
  summaryFunction = twoClassSummary)  # Use metrics like ROC, Sensitivity, Specificity

```

**method = "repeatedcv":** This tells caret::train() to use repeated k-fold cross-validation.

**number = 10:** Specifies 10 folds (i.e., 10-fold cross-validation).

**repeats = 5:** The entire 10-fold process will be repeated 5 times, each with a different random split.

**This setup is useful for:**

   - Reducing variance in the performance estimate.

   - Ensuring the model's performance is stable across different splits of the data.
   
   - Getting a more reliable estimate of the model’s accuracy, precision, RMSE, or other metrics.

# Train Machine learning Model

# Logistic Regression

```{r}
# Load libraries
library(caret)               ## for training machine learning models
library(tictoc)              ## for determining the time taken for a model to run

tic()
set.seed(123)                ## Ensures that results are reproducible

LRModel <- train(factor(Malaria_Result) ~ ., 
                 data = over, 
                 method = "glm", 
                 trControl = control)
toc()
LRModel

```

```{r}
# Prediction using Logistic Regression model
LRpred = predict(LRModel,newdata = test)

# Evaluation of the Logistic Regression model performance metrics
LR_CM <- confusionMatrix(LRpred,as.factor(test$Malaria_Result), positive = "Positive", mode='everything')
LR_CM

```

```{r}
# Combine data into a data frame
Ground_truth<- test$Malaria_Result
Predicted <- LRpred

resultLR <- data.frame(Ground_truth, Predicted)
resultLR$Correct <- resultLR$Ground_truth == resultLR$Predicted

# Add a column for classification results (correct/incorrect)
resultLR<- data.frame(test, LRpred, resultLR$Correct)
#print(resultLR)

```


## Plot of Confusion Matrix of Logistic Regression
```{r}
#load required packages
library(reshape2)
library(scales)

# Create the confusion matrix
conf_matrix <- matrix(c(194, 11, 2, 43), nrow = 2, byrow = TRUE)

# Name the rows and columns
rownames(conf_matrix) <- c("Negative","Positive")
colnames(conf_matrix) <- c("Negative","Positive")

# Melt the matrix for ggplot
conf_df <- melt(conf_matrix)

# Plot
ggplot(conf_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "red", size = 6) +
  scale_fill_gradient(low = "white", high = "purple") +
  labs(title = "Logistic Regreession Confusion Matrix", x = "Predicted label", y = "True label") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text = element_text(face = "bold"))

```
## Model performance metric for classification
**Confusion matrix **: is a performance evaluation tool used for classification models. It shows how well the model's predicted classes match the actual classes.

 <div style='text-align:center;margin:auto;display:block:'><IMG  src=  "C:\Users\Admin\Documents\2025_CDAM_WORKSHOP_1\confusion-matrix.jpeg"></div>
 
********************************************************************************
**Accuracy: The proportion of all predictions that are correct ** used to determine the performance of the Algorithms

$$
\text{Accuracy} = \frac{TP + TN}{TP + FP + FN + TN} =\frac{43 + 194}{43 + 11 + 2 + 194} = 0.948
$$ 

**Interpretation:**

   - How often the classifier is correct overall.
   
   -⚠️ Can be misleading if classes are imbalanced.
   
********************************************************************************
**True Positive Rate (TPR): Also known as sensitivity or recall** 

   - It is the ratio of correctly predicted positive observations to the actual positives. 

$$
\text{Sensitivity} = \frac{TP}{TP + FN} = 0.956
$$

where TP is True Positives and FN is False Negatives.

**Interpretation:**

   - Of all the true positive cases, how many did the model catch?
  
   - 📌 Important in medical diagnosis, fraud detection.
   
********************************************************************************
**False Positive Rate (FPR). Also known as specificity **

   - It is the ratio of incorrectly predicted positive observations to the actual negatives.
   
$$
\text{Specificity} = \frac{TN}{TN + FP} = 0.946
$$
 

where FP is False Positives and TN is True Negatives.

**Interpretation:**

   - Of all the true negative cases, how many did the model correctly identify?

   - 📌 Important to minimize false alarms.
   

********************************************************************************
**Precision: The proportion of predicted positives that are actually positive**

used to test the correctness of the model when it gives a positive outcome

$$
\text{Precision} = \frac{TP}{TP + FP} = 0.796
$$

**Interpretation:**

  - How trustworthy are the positive predictions?

  -📌 Important when false positives are costly (e.g., spam filters, cancer screening, malaria detection).

********************************************************************************

**F1 score: The harmonic mean of precision and recall. It balances the two when both are important.**

Used to assess the number of variables that are missing in the predictions which are positive target

$$
F_1 Score = \frac{2 \cdot \text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}} = 0.869
$$

**Interpretation:**

   - F1 reaches its best value at 1 (perfect precision and recall) and worst at 0.

   - 📌 Useful in imbalanced classification tasks.

********************************************************************************

## Importance of features in a Logistic regression Model

```{R}
# Show relative importance of features
# vip::vip(LRModel)

# Alternatively using ggplot function
var_imp <-varImp(LRModel)
ggplot(var_imp, aes(x = reorder(Variable, Importance), y = importance)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +
  xlab("Variable") +
  ylab("Importance") +
  ggtitle("Feature Importance Plot for LR Model") +
  theme(plot.title = element_text(hjust = 0.5)) # Align the title to the center

```

## Prepare data for explain() function

```{R}
library(DALEX)  ## powerful tool for interpreting machine learning models.

# Converts the target variable from categorical ("Positive"/"Negative") to numeric (1=for Positive, 0=otherwise) 
# Required for many ML models and for DALEX to interpret the output as binary classification.

train$Malaria_Result <- ifelse(train$Malaria_Result == "Positive", 1, 0) 

# Create the explainer Object
explainer_1 <- explain(model = LRModel,
                     data = train[, -which(names(train) =="Malaria_Result")],  # Exclude the target column
                     y = train$Malaria_Result,                                 # Target values as vector
                     label = "Local Explanation with DALEX for Logistics Regression")

# This creates an explainer1 object used by the DALEX and ingredients packages for interpretability methods
```

```{r}
# Select an instance to explain from test set or new unseen data 

set.seed(123)
new_observation_1 <- test[3, -which(names(test)=="Malaria_Result")]  # Select a test instance
#new_observation_1

# Break Down explanation for the instance
local_explanation_1 <- predict_parts(explainer_1, new_observation_1)

# Plot local explanation
plot(local_explanation_1)

```

## Overview
The graph presents a local explanation of a RF model using the DALEX package. It visualizes
how different predictor variables contribute to the prediction for a specific instance. The prediction is
represented by the bar on the right, and the contributions of each variable are shown as horizontal bars.

## Breakdown of Contributions
• Intercept: This baseline value represents the model’s prediction when all predictor variables are zero
 or absent. In this case, the intercept is xxxxxx
 
• Predictor Variables: Each predictor variable’s contribution is shown as a bar. The color indicates the
 direction of the contribution:
 
• Green: Positive contribution, meaning the variable increases the prediction.

• Red: Negative contribution, meaning the variable decreases the prediction.

• The length of the bar represents the magnitude of the contribution.

## Overall Prediction

Summing up all the contributions (intercept + predictor variables), we arrive at the final prediction of xxxx
This value represents the probability of a positive outcome, as RF model typically output probabilities.

This graph provides a valuable tool for understanding how a RF model arrives at a specific prediction. It highlights the relative importance of different predictor variables and their impact on the final outcome.

However, it is essential to consider the limitations and interpret the results in conjunction with other model evaluation metrics

## SHAP(SHapley Additive exPlanations)

The Shapley value helps explain how much each feature contributes to the prediction made by a machine
learning model. It provides a way to fairly distribute the “credit” for the model’s output across all input
features. By visualizing the SHAP plot, you can understand not only which features are important, but also
how specific feature values that are driving predictions for individual cases.

## SHAP explanation

```{r}
shap_values_1 <- predict_parts(explainer_1, new_observation_1, type = "shap")

plot(shap_values_1)

```

## Overview 

1. Leftward(negative): Indicates the feature is pushing the model prediction towards a negative class

2. Larger absolute SHAP values mean a feature has a stronger influence on the prediction.

3. Smaller SHAP values (close to zero) indicate that a feature has minimal influence on the model’s output for
that instance

## Partial Dependence Plots (PDP)

PDP show the relationship between a feature and the predicted outcome while keeping other features constant. This can be particularly useful in diabetes prediction models, as it allows clinicians to visualize how changes in a single factor (e.g., glucose levels or BMI) affect the likelihood of diabetes while accounting for other variables. 

  - Understand the relationship between features and predictions.
    
  - Visualizes the relationship between features and the target prediction
    
  - Understand non-linear relationships

```{r}
## Partial Dependence Plot for Temperature

pdp <- model_profile(explainer_1, variables = "Temperature", type = "partial")
plot(pdp)

## Partial Dependence Plot for Humidity

pdp <- model_profile(explainer_1, variables = "Humidity", type = "partial")
plot(pdp)

```

# Random Forest 

This is an ensemble learning method that combines multiple decision trees to improve prediction accuracy and reduce variance.
## mtry
This parameter controls the number of features randomly chosen as candidates for splitting a node in each tree.
## Training the RF model
```{R}
# Load libraries
library(caret)

set.seed(123)
tuneGrid_rf <- expand.grid(mtry = c(2, 4, 6, 8, 12))

tic()
RFModel <- train(factor(Malaria_Result)~., 
                 data=over, 
                 method="rf", 
                 trControl=control, 
                 tuneGrid=tuneGrid_rf)
toc()
RFModel
#plot(RFModel)
```

## Prediction & Evaluation of the RF model performance metrics

```{R}
# Prediction using RF model
RFpred=predict(RFModel,newdata = test)

# Evaluation of the RF model performance metrics
RF_CM<- confusionMatrix(RFpred,as.factor(test$Malaria_Result), positive = "Positive", mode='everything')
RF_CM

```

## Plot of Confusion Matrix of Random Forest
```{r}
#load required packages
library(reshape2)
library(scales)

# Create the confusion matrix
conf_matrix <- matrix(c(201, 5, 4, 40), nrow = 2, byrow = TRUE)

# Name the rows and columns
rownames(conf_matrix) <- c("Negative","Positive")
colnames(conf_matrix) <- c("Negative","Positive")

# Melt the matrix for ggplot
conf_df <- melt(conf_matrix)

# Plot
ggplot(conf_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "black", size = 6) +
  scale_fill_gradient(low = "white", high = "purple") +
  labs(title = "Random Forest Confusion Matrix", x = "Predicted label", y = "True label") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text = element_text(face = "bold"))
```

## Importance of features in a Random Model

```{R}
# Show relative importance of features
# vip::vip(RFModel)

# Alternatively using ggplot function
var_imp <-varImp(RFModel)
ggplot(var_imp, aes(x = reorder(Variable, Importance), y = importance)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +
  xlab("Variable") +
  ylab("Importance") +
  ggtitle("Feature Importance Plot for RF Model") +
  theme(plot.title = element_text(hjust = 0.5)) # Align the title to the center

```

## Prepare data for explain() function
```{R}
library(DALEX)

# Converts the target variable from categorical ("Positive"/"Negative") to numeric (1 for Positive, 0 otherwise) 
# Required for many ML models and for DALEX to interpret the output as binary classification.

train$Malaria_Result <- ifelse(train$Malaria_Result == "Positive", 1, 0) 

# Create the explainer Object
explainer_2 <- explain(model = RFModel,
                     data = train[, -which(names(train) =="Malaria_Result")],  # Exclude the target column
                     y = train$Malaria_Result,                                 # Target values as vector
                     label = "Local Explanation with DALEX for Random Forest")

# This creates an explainer object used by the DALEX and ingredients packages for interpretability methods
```


```{r}
# Select an instance to explain
new_observation_2 <- test[3, -which(names(test)=="Malaria_Result")]  # Select a test instance
new_observation_2

# Break Down explanation for the instance
local_explanation_2 <- predict_parts(explainer_2, new_observation_2)

# Plot local explanation
plot(local_explanation_2)

```
## Overview

The graph presents a local explanation of a RF model using the DALEX package. It visualizes
how different predictor variables contribute to the prediction for a specific instance. The prediction is
represented by the bar on the right, and the contributions of each variable are shown as horizontal bars.

## Breakdown of Contributions

• Intercept: This baseline value represents the model’s prediction when all predictor variables are zero
 or absent. In this case, the intercept is 0.221
 
• Predictor Variables: Each predictor variable’s contribution is shown as a bar. The color indicates the
 direction of the contribution:
 
• Green: Positive contribution, meaning the variable increases the prediction.

• Red: Negative contribution, meaning the variable decreases the prediction.

• The length of the bar represents the magnitude of the contribution.

## Overall Prediction
Summing up all the contributions (intercept + predictor variables), we arrive at the final prediction of 0.95.
This value represents the probability of a positive outcome, as RF model typically output probabilities.

This graph provides a valuable tool for understanding how a RF model arrives at a specific prediction. It highlights the relative importance of different predictor variables and their impact on the final outcome.

However, it is essential to consider the limitations and interpret the results in conjunction with other model evaluation metrics
 
## SHAP(SHapley Additive exPlanations)

The Shapley value helps explain how much each feature contributes to the prediction made by a machine
learning model. It provides a way to fairly distribute the “credit” for the model’s output across all input
features. By visualizing the SHAP plot, you can understand not only which features are important, but also
how specific feature values that are driving predictions for individual cases.

## SHAP explanation

```{r}
shap_values_2 <- predict_parts(explainer_2, new_observation_2, type = "shap")

plot(shap_values_2)

```

## Overview 

1. Leftward(negative): Indicates the feature is pushing the model prediction towards a negative class

2. Larger absolute SHAP values mean a feature has a stronger influence on the prediction.

3. Smaller SHAP values (close to zero) indicate that a feature has minimal influence on the model’s output for
that instance

## Partial Dependence Plots (PDP)

PDP show the relationship between a feature and the predicted outcome while keeping other features constant. This can be particularly useful in diabetes prediction models, as it allows clinicians to visualize how changes in a single factor (e.g., glucose levels or BMI) affect the likelihood of diabetes while accounting for other variables. 

  - Understand the relationship between features and predictions.
    
  - Visualizes the relationship between features and the target prediction
    
  - Understand non-linear relationships

```{r}
## Partial Dependence Plot for Humidity

pdp <- model_profile(explainer_2, variables = "Humidity", type = "partial")
plot(pdp)

```

## Support Vector Machine (SVM)
```{r}
#load library
library(caret)

#Tune the grid
svm_tunegrid <- expand.grid(sigma = c(0.01, 0.1, 0.2), C = c(0.1, 1, 10))

tic()
set.seed(123)                ## Ensures that results are reproducible

# Training a model with standardization: This step is useful for models that are sensitive to the scale of the data such as:KNN, SVM, DT etc
SVMModel <- train(factor(Malaria_Result) ~ ., 
                  data = over, 
                  method = "svmRadial", 
                  trControl = control,
                  tuneGrid =svm_tunegrid,
                  preProcess= c("center", "scale"))      ##function to normalize the predictors

toc()

```

## View the Model
```{r}
SVMModel
```

## View the Best Tune
```{r}
SVMModel$bestTune
```

## View the Results
```{r}
SVMModel$results
```

## Plot the Best Model
```{r}
plot(SVMModel)
```

## Prediction & Evaluation of the SVM model performance metrics
```{r}
SVMpred = predict(SVMModel, newdata = test)
SVM_CM <- confusionMatrix(SVMpred, as.factor(test$Malaria_Result), positive = "Positive", mode='everything')
SVM_CM
```

## Plot of Confusion Matrix of SVM
```{r}
#load required packages
library(reshape2)
library(scales)
# Create the confusion matrix
conf_matrix <- matrix(c(200, 5, 4, 41), nrow = 2, byrow = TRUE)
# Name the rows and columns
rownames(conf_matrix) <- c("Negative","Positive")
colnames(conf_matrix) <- c("Negative","Positive")
# Melt the matrix for ggplot
conf_df <- melt(conf_matrix)
# Plot
ggplot(conf_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "black", size = 6) +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "SVM Confusion Matrix", x = "Predicted label", y = "True label") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text = element_text(face = "bold")
  )
```

## Prepare data for explain() function
```{r}
# Converts the target variable from categorical ("Positive"/"Negative") to numeric (1 for Positive, 0 otherwise)
train$Malaria_Result <- ifelse(train$Malaria_Result == "Positive", 1, 0)
# Create the explainer Object
explainer_3 <- explain(model = SVMModel,
                     data = train[, -which(names(train) =="Malaria_Result")],  # Exclude the target column
                     y = train$Malaria_Result,                                 # Target values as vector
                     label = "Local Explanation with DALEX for SVM")
```

This creates an explainer object used by the DALEX and ingredients packages for interpretability methods

```{r}
new_observation_3 <- test[3, -which(names(test)=="Malaria_Result")]  # Select a test instance
new_observation_3

# Break Down explanation for the instance
local_explanation_3 <- predict_parts(explainer_3, new_observation_3)

# Plot local explanation
plot(local_explanation_2)
```
## Overview

The graph presents a local explanation of a RF model using the DALEX package. It visualizes
how different predictor variables contribute to the prediction for a specific instance. The prediction is
represented by the bar on the right, and the contributions of each variable are shown as horizontal bars.

## Breakdown of Contributions
• Intercept: This baseline value represents the model’s prediction when all predictor variables are zero
 or absent. In this case, the intercept is 0.221
 
• Predictor Variables: Each predictor variable’s contribution is shown as a bar. The color indicates the
 direction of the contribution:
 
• Green: Positive contribution, meaning the variable increases the prediction.

• Red: Negative contribution, meaning the variable decreases the prediction.

• The length of the bar represents the magnitude of the contribution.

## Overall Prediction
Summing up all the contributions (intercept + predictor variables), we arrive at the final prediction of 0.95.
This value represents the probability of a positive outcome, as RF model typically output probabilities.

This graph provides a valuable tool for understanding how a RF model arrives at a specific prediction. It highlights the relative importance of different predictor variables and their impact on the final outcome.

However, it is essential to consider the limitations and interpret the results in conjunction with other model evaluation metrics
 
## SHAP(SHapley Additive exPlanations)
The Shapley value helps explain how much each feature contributes to the prediction made by a machine
learning model. It provides a way to fairly distribute the “credit” for the model’s output across all input
features. By visualizing the SHAP plot, you can understand not only which features are important, but also
how specific feature values that are driving predictions for individual cases.

## SHAP explanation

```{r}
shap_values_3 <- predict_parts(explainer_3, new_observation_3, type = "shap")

plot(shap_values_3)
```

## Overview 

1. Leftward(negative): Indicates the feature is pushing the model prediction towards a negative class

2. Larger absolute SHAP values mean a feature has a stronger influence on the prediction.

3. Smaller SHAP values (close to zero) indicate that a feature has minimal influence on the model’s output for
that instance

# k-Nearest Neighbors (KNN)
```{r}

tic()
set.seed(123)                ## Ensures that results are reproducible

# Training a model with standardization: This step is useful for models that are sensitive to the scale of the data such as:KNN, SVM, DT etc
KNNModel <- train(factor(Malaria_Result) ~ ., 
                  data = over, 
                  method = "knn", 
                  trControl = control,
                  preProcess= c("center", "scale")) ##function to normalize the predictors
toc()
KNNModel

```

## View the Best Tune
```{r}
KNNModel$bestTune
```

## Plot the Model
```{r}
plot(KNNModel)
```

## Prediction & Evaluation of the KNN model performance metrics
```{r}
KNNpred = predict(KNNModel, newdata = test)
KNN_CM <- confusionMatrix(KNNpred, as.factor(test$Malaria_Result), positive = "Positive", mode='everything')
KNN_CM
```
## Plot of Confusion Matrix of KNN
```{r}
#load required packages
library(reshape2)
library(scales)
# Create the confusion matrix
conf_matrix <- matrix(c(183, 1, 22, 44), nrow = 2, byrow = TRUE)
# Name the rows and columns
rownames(conf_matrix) <- c("Negative","Positive")
colnames(conf_matrix) <- c("Negative","Positive")
# Melt the matrix for ggplot
conf_df <- melt(conf_matrix)
# Plot
ggplot(conf_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "black", size = 6) +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "KNN Confusion Matrix", x = "Predicted label", y = "True label") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text = element_text(face = "bold")
  )
```

## Prepare data for explain() function
```{r}
# Converts the target variable from categorical ("Positive"/"Negative") to numeric (1 for Positive, 0 otherwise)
train$Malaria_Result <- ifelse(train$Malaria_Result == "Positive", 1, 0)
# Create the explainer Object
explainer_4 <- explain(model = KNNModel,
                     data = train[, -which(names(train) =="Malaria_Result")],  # Exclude the target column
                     y = train$Malaria_Result,                                 # Target values as vector
                     label = "Local Explanation with DALEX for KNN")
```

This creates an explainer object used by the DALEX and ingredients packages for interpretability methods

```{r}
new_observation_4 <- test[3, -which(names(test)=="Malaria_Result")]  # Select a test instance
new_observation_4

# Break Down explanation for the instance
local_explanation_4 <- predict_parts(explainer_4, new_observation_4)
# Plot local explanation
plot(local_explanation_4)
```

## Overview
The graph presents a local explanation of a KNN model using the DALEX package. It visualizes how different predictor variables contribute to the prediction for a specific instance. The prediction is represented by the bar on the right, and the contributions of each variable are shown as horizontal bars.

## SHAP explanation

```{r}
shap_values_4 <- predict_parts(explainer_4, new_observation_4, type = "shap")

plot(shap_values_4)
```

## Decision Tree (DT)
```{r}
set.seed(123)                ## Ensures that results are reproducible
tic()
DTModel <- train(factor(Malaria_Result) ~ ., 
                  data = over, 
                  method = "rpart", 
                  trControl = control,
                  preProcess= c("center", "scale"))
toc()
```

## View the Model
```{r}
DTModel
```

## View the Best Tune
```{r}
DTModel$bestTune
```

## Plot the Model
```{r}
plot(DTModel)
```

## Fancy Rpartplot
```{r}
library(rpart.plot)
rpart.plot(DTModel$finalModel, 
           main = "Decision Tree for Malaria Prediction",
           extra = 104, 
           type = 3, 
           fallen.leaves = TRUE, 
           box.palette = "RdBu", 
           shadow.col = "gray", 
           nn = TRUE)
```

## Prediction & Evaluation of the DT model performance metrics
```{r}
DTpred = predict(DTModel, newdata = test)
DT_CM <- confusionMatrix(DTpred, as.factor(test$Malaria_Result), positive = "Positive", mode='everything')
DT_CM
```

## Plot of Confusion Matrix of DT
```{r}
#load required packages
library(reshape2)
library(scales)
# Create the confusion matrix
conf_matrix <- matrix(c(175, 6, 30, 39), nrow = 2, byrow = TRUE)
# Name the rows and columns
rownames(conf_matrix) <- c("Negative","Positive")
colnames(conf_matrix) <- c("Negative","Positive")
# Melt the matrix for ggplot
conf_df <- melt(conf_matrix)
# Plot
ggplot(conf_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "black", size = 6) +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Decision Tree Confusion Matrix", x = "Predicted label", y = "True label") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text = element_text(face = "bold")
  )
```

## Prepare data for explain() function
```{r}
# Converts the target variable from categorical ("Positive"/"Negative") to numeric (1 for Positive, 0 otherwise)
train$Malaria_Result <- ifelse(train$Malaria_Result == "Positive", 1, 0)
# Create the explainer Object
explainer_5 <- explain(model = DTModel,
                     data = train[, -which(names(train) =="Malaria_Result")],  # Exclude the target column
                     y = train$Malaria_Result,                                 # Target values as vector
                     label = "Local Explanation with DALEX for Decision Tree")
```

```{r}
# This creates an explainer object used by the DALEX and ingredients packages for interpretability methods
new_observation_5 <- test[3, -which(names(test)=="Malaria_Result")]  # Select a test instance
new_observation_5
# Break Down explanation for the instance
local_explanation_5 <- predict_parts(explainer_5, new_observation_5)
# Plot local explanation
plot(local_explanation_5)
```

## SHAP explanation

```{r}
shap_values_5 <- predict_parts(explainer_5, new_observation_5, type = "shap")

plot(shap_values_5)
```

## Naive Bayes NB
```{r}
set.seed(123)                ## Ensures that results are reproducible
tic()
NBModel <- train(factor(Malaria_Result) ~ ., 
                  data = over, 
                  method = "naive_bayes", 
                  trControl = control,
                  preProcess= c("center", "scale"))
toc()
```

## View the Model
```{r}
NBModel
```

## View the Best Tune
```{r}
NBModel$bestTune
```

## View the Results
```{r}
NBModel$results
```

## Plot the Model
```{r}
plot(NBModel)
```

## Prediction & Evaluation of the NB model performance metrics
```{r}
NBpred = predict(NBModel, newdata = test)
NB_CM <- confusionMatrix(NBpred, as.factor(test$Malaria_Result), positive = "Positive", mode='everything')
NB_CM
```

## Plot of Confusion Matrix of NB
```{r}
#load required packages
library(reshape2)
library(scales)
# Create the confusion matrix
conf_matrix <- matrix(c(188, 1, 17, 44), nrow = 2, byrow = TRUE)
# Name the rows and columns
rownames(conf_matrix) <- c("Negative","Positive")
colnames(conf_matrix) <- c("Negative","Positive")
# Melt the matrix for ggplot
conf_df <- melt(conf_matrix)
# Plot
ggplot(conf_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "black", size = 6) +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Naive Bayes Confusion Matrix", x = "Predicted label", y = "True label") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text = element_text(face = "bold")
  )
```

## Prepare data for explain() function
```{r}
# Converts the target variable from categorical ("Positive"/"Negative") to numeric (1 for Positive, 0 otherwise)
train$Malaria_Result <- ifelse(train$Malaria_Result == "Positive", 1, 0)
# Create the explainer Object
explainer_6 <- explain(model = NBModel,
                     data = train[, -which(names(train) =="Malaria_Result")],  # Exclude the target column
                     y = train$Malaria_Result,                                 # Target values as vector
                     label = "Local Explanation with DALEX for Naive Bayes")
```

This creates an explainer object used by the DALEX and ingredients packages for interpretability methods

```{r}
new_observation_6 <- test[2, -which(names(test)=="Malaria_Result")]  # Select a test instance
new_observation_6

# Break Down explanation for the instance
local_explanation_6 <- predict_parts(explainer_6, new_observation_6)
# Plot local explanation
plot(local_explanation_6)
```

### SHAP explanation
```{r}
shap_values_6 <- predict_parts(explainer_6, new_observation_6, type = "shap")
plot(shap_values_6)
```

## ASSIGNMENT(Try for other Models)
# NB: You can try other models like XGBoost, LightGBM, CatBoost, etc. by following the same pattern as above.

```{r}

```

